---
title: "Comparison of DTU"
author: "Christoph Kiefer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        toc_depth: 2
        fig_caption: false
        dev: "svg"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
    cache = FALSE, out.width = 800)
```

```{r library, include=FALSE}
source(".Rprofile")
suppressPackageStartupMessages({
    library("dplyr")
    library("readr")
    library("purrr")
    library("UpSetR")
    library("ggplot2")
    library("ggthemes")
    library("topGO")
    library("org.Mm.eg.db")
    library("IsoformSwitchAnalyzeR")
    source("R/go.R")
    source("R/reactomePA.R")
})
```

```{r dataImport}
df <- snakemake@input$dtu_res %>% 
    set_names(basename(.) %>% 
                  sub("_dtu.+$", "", .)) %>% 
    map(read_csv, col_types = c("cccfdddd"))

# add ensembl gene_id
is <- read_tsv(snakemake@input$illu_tmap) %>% 
    dplyr::select(qry_gene_id, ref_gene_id) %>% 
    tibble::deframe()

df$illumina_illumina_stringtie <- df$illumina_illumina_stringtie %>% 
    mutate(gene_id = is[gene_id])

df$cdna_illumina_stringtie <- df$cdna_illumina_stringtie %>% 
    mutate(gene_id = is[gene_id])


ts <- read_tsv(snakemake@input$telo_tmap) %>% 
    dplyr::select(qry_gene_id, ref_gene_id) %>% 
    tibble::deframe()

df$illumina_teloprime_stringtie <- df$illumina_teloprime_stringtie %>% 
    mutate(gene_id = ts[gene_id])

df$cdna_teloprime_stringtie <- df$cdna_teloprime_stringtie %>% 
    mutate(gene_id = ts[gene_id])


annotation <- loadDb(snakemake@input$annotation)

df$illumina_ref <- df$illumina_ref %>% 
    mutate(gene_id = mapIds(
        keys = .$transcript_id,
        x = annotation,
        column = "GENEID",
        keytype = "TXNAME",
        multiVals = "first"))

df$cdna_ref <- df$cdna_ref %>% 
    mutate(gene_id = mapIds(
        keys = .$transcript_id,
        x = annotation,
        column = "GENEID",
        keytype = "TXNAME",
        multiVals = "first"))


# bind
df <- df %>% 
    bind_rows(.id = "id")


# import swithc analyzer lists and extract switching type info
switch_conseq <- snakemake@input$sal %>% 
    set_names(basename(.) %>% 
                  sub("_sal.rds$", "", .)) %>% 
    map(read_rds) %>% 
    map(extractSplicingSummary,
        alpha = snakemake@params$dtu_cutoff,
        dIFcutoff = snakemake@params$dIF_cutoff,
        plot = FALSE,
        returnResult = TRUE) %>% 
    map(as_tibble) %>% 
    bind_rows(.id = "dataset") %>% 
    group_by(dataset, AStype) %>% 
    summarise(nrGenesWithConsequences = sum(nrGenesWithConsequences))
```

# Gene level overlap

only genes in reference

## Enrichment analysis

### GO

```{r}
# list of reference vectors
reference <- df %>% 
    dplyr::select(id, gene_id, gene) %>% 
    group_by(id, gene_id) %>% 
    summarise(gene = min(gene)) %>% 
    filter(grepl("^ENS", gene_id)) %>% 
    split(~id) %>% 
    map(pull, gene_id) %>% 
    map(remove.version)
```


Alle Gene mit p < .1. Referenz/universum = Alle Gene, die die entsprechenden Filter
in DrimSeq überstanden haben (also für jeden Datensatz unterschiedlich.)

```{r, fig.asp = .65}
go_res <- df %>% 
    dplyr::select(id, gene_id, gene) %>% 
    filter(gene < snakemake@params$dtu_cutoff) %>% 
    split(~id) %>% 
    map(pull, "gene_id") %>% 
    map(remove.version) %>% 
    map2(reference, get_geneList) %>% 
    map2(names(.), make.topGO) %>% 
    map(get.results, method = "parentchild")

plot.go.summary(go_res, n = 3, n_char = 50) +
    xlab(NULL) +
    theme_tufte(base_size = 8, base_family = "Helvetica") +
    geom_rangeframe() +
    scale_fill_viridis_c(direction = -1, trans = "log10", name = "p") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

100 genes with smallest p.

```{r, fig.asp = .65}
go_res <- df %>% 
    dplyr::select(id, gene_id, gene) %>% 
    group_by(id) %>% 
    slice_min(order_by = gene, n = 100) %>% 
    ungroup() %>% 
    split(~id) %>% 
    map(pull, "gene_id") %>% 
    map(remove.version) %>% 
    map2(reference, get_geneList) %>% 
    map2(names(.), make.topGO) %>% 
    map(get.results, method = "parentchild")

plot.go.summary(go_res, n = 3, n_char = 50) +
    xlab(NULL) +
    ggplot2::scale_fill_viridis_c(direction = -1, trans = "log10", name = "p") +
    theme_tufte(base_size = 8, base_family = "Helvetica") +
    geom_rangeframe() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Reactome

No significant terms.

```{r}
# named vector of entrez_ids
reference <- df %>% 
    dplyr::select(id, gene_id, gene) %>% 
    group_by(id, gene_id) %>% 
    summarise(gene = min(gene)) %>% 
    filter(grepl("^ENS", gene_id)) %>% 
    split(~id) %>% 
    map(pull, gene_id) %>% 
    map(remove.version) %>% 
    map(mapIds,
        x = org.Mm.eg.db,
        column = "ENTREZID",
        keytype = "ENSEMBL",
        multiVals = "first") %>% 
    map(unlist)

reactome_res <- df %>% 
    dplyr::select(id, gene_id, gene) %>% 
    filter(gene < snakemake@params$dtu_cutoff) %>% 
    split(~id) %>% 
    set_names(sub("svalue_", "", names(.))) %>% 
    map(pull, gene_id) %>% 
    map(remove.version) %>% 
    map(mapIds,
        x = org.Mm.eg.db,
        column = "ENTREZID",
        keytype = "ENSEMBL",
        multiVals = "first") %>% 
    map2(reference, ReactomePA::enrichPathway,
        organism = "mouse",
        pvalueCutoff = 0.05,
        pAdjustMethod = "BH",
        qvalueCutoff = 0.2,
        readable = TRUE)
```

```{r, fig.asp = .65}
plot.reactome.summary(reactome_res)
#    xlab(NULL) +
#    ggplot2::scale_fill_viridis_c(direction = -1, trans = "log10", name = "q") +
#    theme_tufte(base_size = 8, base_family = "Helvetica") +
#    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Upset

```{r, fig.asp = .65}
df %>%
    filter(grepl("^ENS", gene_id)) %>% 
    group_by(id, gene_id) %>% 
    summarise(gene = min(gene)) %>% 
    ungroup() %>% 
    mutate(gene = tidyr::replace_na(gene, 1)) %>% 
    mutate(gene = as.integer(gene <= snakemake@params$dtu_cutoff)) %>% 
    tidyr::pivot_wider(values_from = gene, names_from = id, values_fill = 0) %>% 
    as.data.frame() %>%
    tibble::column_to_rownames("gene_id") %>%
    upset(nsets = 8, keep.order = TRUE,
        order.by = "freq",
        sets = c("illumina_ref", "illumina_illumina_stringtie",
                 "illumina_cdna_flair", "illumina_teloprime_stringtie",
                 "cdna_ref", "cdna_illumina_stringtie",
                 "cdna_cdna_flair", "cdna_teloprime_stringtie"),
        sets.x.label = "Genes with DTU",
        mainbar.y.label = "Shared genes with DTU",
        text.scale = 1.25)
```

## Table

```{r}
df %>% 
    group_by(gene_id) %>% 
    mutate(min_gene = min(gene)) %>% 
    ungroup() %>% 
    filter(min_gene < snakemake@params$dtu_cutoff) %>% 
    group_by(id, gene_id, mgi_symbol, gene_biotype) %>% 
    summarise(gene = min(gene)) %>% 
    ungroup() %>% 
    tidyr::pivot_wider(values_from = gene, names_from = id, values_fill = 1) %>% 
    filter(grepl("^ENS", gene_id)) %>% 
    arrange(mgi_symbol) %>% 
    knitr::kable(format = "html") %>%
    kableExtra::kable_styling() %>%
    kableExtra::scroll_box(width = "100%", height = "600px")
```

## Splicing type analysis

A single gene can show multiple different kinds of events, therefore the numbers
can be larger then in the upsetplot above. (This is the reason why fig 5d in chen
et al is an upset plot. It shows which events occur together in a single gene.)

```{r, fig.width = 6.85 / 2, fig.asp = .65}
switch_conseq %>% 
    mutate(dataset = factor(dataset, levels = c("illumina_ref", "illumina_illumina_stringtie",
                 "illumina_cdna_flair", "illumina_teloprime_stringtie",
                 "cdna_ref", "cdna_illumina_stringtie",
                 "cdna_cdna_flair", "cdna_teloprime_stringtie"))) %>%
    ggplot(aes(dataset, nrGenesWithConsequences)) +
    geom_bar(aes(fill = AStype), stat = "identity") +
    scale_fill_brewer(palette = "Dark2", name = NULL) +
    theme_tufte(base_size = 8, base_family = "Helvetica") +
    geom_rangeframe(sides = 'bl') +
    ylab("Isoform switching events") +
    xlab(NULL) +
    theme(legend.key.size = unit(.33, "cm")) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```


# qPCR validation

```{r, fig.width = 6.85 / 2, fig.asp = .65}
qpcr <- read_csv(snakemake@input$qpcr, show_col_types = FALSE)

df %>%
    group_by(id, gene_id) %>% 
    summarise(gene = min(gene)) %>% 
    ungroup() %>% 
    tidyr::pivot_wider(values_from = gene, names_from = id, values_fill = 1) %>% 
    mutate(mgi_symbol = mapIds(
        keys = remove.version(gene_id),
        x = org.Mm.eg.db,
        column = "SYMBOL",
        keytype = "ENSEMBL",
        multiVals = "first")) %>% 
    right_join(qpcr, by = "mgi_symbol") %>% 
    dplyr::select(-qpcr_p, -sign, -gene_id) %>% 
    tidyr::gather("dataset", "padj", -mgi_symbol) %>%  
    mutate(dataset = if_else(dataset == "qpcr_padj", "qpcr", dataset)) %>% 
    mutate(mgi_symbol = factor(mgi_symbol, levels = rev(levels(as.factor(mgi_symbol))))) %>% 
    filter(!mgi_symbol %in% c("Smyd4", "Pex6", "Lipe", "Gnas")) %>% 
    mutate(dataset = factor(dataset, levels = c("illumina_ref", "illumina_illumina_stringtie",
                 "illumina_cdna_flair", "illumina_teloprime_stringtie",
                 "cdna_ref", "cdna_illumina_stringtie",
                 "cdna_cdna_flair", "cdna_teloprime_stringtie", "qpcr"))) %>% 
    ggplot(aes(dataset, mgi_symbol)) +
        geom_tile(aes(fill = padj)) +
        scale_fill_viridis_c(direction = -1, labels = scales::label_pvalue(),
            limits = c(0,1)) +
        theme_tufte(base_size = 8, base_family = "Helvetica") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
        xlab(NULL) + 
        scale_y_discrete(name = NULL) +
        theme(legend.key.size = unit(.33, "cm"))
```

```{r}
qpcr %>% 
    filter(!mgi_symbol %in% c("Smyd4", "Pex6", "Lipe", "Gnas")) %>%  
    mutate(qpcr_padj = p.adjust(qpcr_p)) %>%
    mutate(sign = symnum(qpcr_padj,
                         cutpoints = c(0, .001, .01, .05, .1, 1),
                         symbols = c("***", "**", "*", ".", " ")))  %>% 
      knitr::kable(format = "html") %>%
    kableExtra::kable_styling()
```
